---
title: "The PROFOUND R-package: ProfoundData"
author: "Ramiro Silveyra Gonzalez, Christopher Reyer, Florian Hartig"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Describes the PROFOUND R-package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
--- 
  

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=10, warning=FALSE, cache = F)
```


```{r setup, echo=F, warning = F, message=F, results = 'hide'}
set.seed(1)
library(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
#panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
#set.alignment('left', row.names = 'right')
library(knitr)
library(plyr)

set.seed(1)


``` 



# ProfoundData R-package: using the PROFOUND database from R 

The ProfoundData package serves as interface to access the [PROFOUND database](PROFOUNDdatabase.html). Three main functions are included to achieve this goal: a browseData function for exploring the database, a getData function for downloading data fromm the database and a plotData function to quickly inspect any variable of the datasets.

Currently, the functions allow to download data from one site and one dataset at a time. For downloading data of more than one dataset or of more than one site, the function should be called as many times as the number of desired sites or datasets. Additionally, flags and other parameters can be passed to the functions.

The package also includes utilities functions to do tasks such as inspecting the data or writing NetCDF files. 

To cite the package, run

```{r}
library(ProfoundData)
citation("ProfoundData")
```

# First Steps

## Database connection

After loading ProfoundData, the first step is to call the *setDB*. This function will create a database object. It requires from you a **valid absolute path to the PROFOUND database**.

```{r, eval = T}
myDB <- path.expand("~/ownCloud/PROFOUND_Data/v0.2.0/ProfoundData.sqlite")
setDB(myDB)
```

To check the database connection you can use the use *getDB*, which returns the database path and also informs about the database version.

```{r, eval = T}
getDB()
```

# Exploring the database

## The browseData function

The browseData function allows to see what data is included in the database and for what sites the data is available. 

```{r, fig.width= 7}
overview <- browseData()
```
```{r , fig.width= 7}
knitr::kable(overview, align = "l")
```

Hint: If you set *collapse* to FALSE, you will get the full version of the overview table.

```{r kable, fig.width= 7, eval = F}
overview <- browseData(collapse = F)
```


Besides returning the overview, browseData allows to check:

+  available datasets

```{r }
tables <- browseData(dataset = "DATASETS")
```

+  available variables for a given dataset

```{r}
variables <- browseData(dataset = "CLIMATE_LOCAL", variables = T)
```

+  available sites for a given dataset

```{r}
available <- browseData(dataset = "CLIMATE_LOCAL")
```


+  available datasets for a given site

```{r}
available <- browseData(site ="le_bray")
```

+  whether a dataset is available for a specific site

```{r}
available <- browseData(site ="le_bray", dataset = "CLIMATE_LOCAL")
```

+  version history 

```{r}
version <- browseData(dataset = "VERSION")
```

+  metadata

```{r}
metadata <- browseData(dataset = "METADATA_CLIMATE_LOCAL")
```
Hint: You can also pass a site to check the metadata of a specific site.
```{r}
metadata <- browseData(dataset = "METADATA_TREE", site = "solling_spruce")
```

+  source

```{r}
source <- browseData(dataset = "SOURCE")
```
Hint: You can also pass a site to check the source of a specific site.
```{r}
source <- browseData(dataset = "SOURCE", site = "solling_spruce")
```

+  policy

```{r}
source <- browseData(dataset = "POLICY")
```
Hint: You can also pass a site to check the policy of a specific site.

```{r}
policy <- browseData(dataset = "POLICY", site = "solling_spruce")
```


## The summarizeData function

### Data overviews of the available data

The *summarizeData* function allows to obtain data summaries and overviews of the data depending on **mode**. If you choose the mode overview, *summarizeData* will return a data frame with the years and number of observations, the first  and last value, and the  min and max value, as in the example below.

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_LOCAL", site = "bily_kriz", mode = "overview")
```
```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

The overview tables for ISIMIP datasets contain more fields, namely forcing datasets and/or forcing conditions. 

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_ISIMIP2B", site = "bily_kriz", mode = "overview")
```
```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

This option is also availabe for TREE and STAND datasets.
```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "TREE", site = "solling_beech", mode = "overview")
```

```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

Please note that the TREE overview table is first summarized by years, pulling all trees together as a stand. Thus, you might find coincidences between the STAND and the TREE overviews.
```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "STAND", site = "solling_beech", mode = "overview")
```

```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

### Data summaries of the available data

The other value of mode is **data**. It sets the behaviour of *summarizeData* to produce result summaries. 

#### TREE


```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "TREE", site = "bily_kriz")
```
```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```


#### CLIMATE

##### CLIMATE_LOCAL

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_LOCAL", site = "bily_kriz")
```
```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

##### CLIMATE_ISIMIP

For several forcing datasets and/or forcing conditions 

```{r, echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_ISIMIP2B", site = "bily_kriz")
```
```{r, echo=T, message= F, warning=F}
kable(data, row.names = F)
```

For only one forcing dataset and forcing conditions 
```{r, echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_ISIMIP2B_IPSLCM5ALR_historical", site = "bily_kriz")
```
```{r, echo=T, message= F, warning=F}
kable(head(data), row.names = F)
```

#### FLUX


```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "FLUX", site = "bily_kriz")
```
```{r, echo=T, message= F, warning=F}
kable(data, row.names = F)
```



# The getData function

The getData allows to download data from the PROFOUND database. The returned object will be a data frame or a list of data frames, depending on the data that has been requested. 


## SITES

To obtain the site information use getData and download the table SITES.

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-getData(dataset = "SITES")
```
```{r, echo=T, message= F, warning=F}
names(data)
```

Optionally, you can specify a site.

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-getData(dataset = "SITES", site = "soro")
```
```{r, echo=T, message= F, warning=F}
knitr::kable(data, row.names = F)
```

A text description can be obtained from **SITEDESCRIPTION**

```{r, fig.show='hold', echo=T, message= F, warning=F}
soro <-getData(dataset = "SITEDESCRIPTION", site = "soro")
```
```{r, echo=T, message= F, warning=F}
soro$description
```


## DATASETS

To download any dataset, you have to provide a dataset name and a site

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-getData( dataset = "CLIMATE_LOCAL", site = "soro")
```


### collapse

This option specifies whether the returned data should be a data frame or a list of data frames. This argument is relevant when downloading data from SOIL or ISIMIP datasets.

For ISIMIP datasets, by setting collapse to FALSE you will obtain a list with data frames named after the forcing datasets and conditions. Otherwise, the data will return in a unique data frame.

```{r, echo=T, message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2A", site = "soro", collapse = FALSE)
```
```{r, echo=T, message= F, warning=F}
names(data)
names(data[[1]])
```

We recommend to unset collapse when downloading SOIL data because in many cases SOIL is a collection of tables.

```{r, echo=T, message= F, warning=F}
data <- getData(dataset ="SOIL", site = "soro", collapse = FALSE)
```
```{r, echo=T, message= F, warning=F}
str(data, 1)
```
```{r, echo=T, message= F, warning=F}
names(data[[1]])
```


### forcingDataset, forcingCondition

With the function arguments *forcingDataset* and *forcingCondition* it is possible to select specific forcing datasets and conditions, respectively. These arguments are relevant for ISIMIP datasets. 

```{r, echo=T, message= F, warning=F}
data <- getData( dataset ="CLIMATE_ISIMIP2B", site ="soro", forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6")
```
```{r, echo=T, message= F, warning=F}
knitr::kable(head(data), align = "l")

```

The previous getData call is actually equivalent to 
```{r, echo=T, message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B_GFDLESM2M_rcp2p6", site ="soro")
```
```{r, echo=T, message= F, warning=F}
knitr::kable(head(data), align = "l")
```


### variables

For any dataset it is possible to define the variable or variables to be downloaded

```{r, echo=T, message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B", site ="soro",
                forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
                variables = "p_mm")
```
```{r, echo=T,  message= F, warning=F}
knitr::kable(head(data),  align = "l")

```

### species
When downloading TREE or STAND data, you can select species with the argument *species*. It takes both full species name or the species id.
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="TREE", site ="hyytiala",
                species = "Pinus sylvestris")
```
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="TREE", site ="hyytiala",
                 species = "pisy")
```

### period
For time series data it is possible to subset the data to a specific period.
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B", site ="soro",
                forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
                period = c("2006-01-01","2006-12-31"))
```
```{r, echo=T, message= F, warning=F}
range(data$date)
```

### quality

Some datasets, such as CLIMATE_LOCAL or FLUX, have quality flags to indicate how the data was obtained. Please be aware of the values of quality flags before using this option. The definition of the flags is available in the metadata. 

When passing a quality flag value, you should also define how the threshold should be applied. Values out of the selected range will be dropped from the data based on your query.

```{r, echo=T, message= F, warning=F}
data <- getData(dataset = "CLIMATE_LOCAL", site = "soro",
                period = c("2011-01-01","2012-12-31"), 
                quality = 1, decreasing = FALSE)

```

```{r, echo=T, message= F, warning=F}
data <- getData(dataset = "FLUX", site = "soro",
                period = c("2011-01-01","2012-12-31"), 
                quality = 0, decreasing = TRUE)
```
# The plotData function
 
The plotData function performs a query to the database, downloads the data, applies the desired options (time period, quality flag, etc.) and shapes the final data into a time series for plotting. It requires a dataset name and a site. Check  the help files to see what datasets are supported. 

## Normal plotting

A simple plot call looks like 

```{r, fig.show='hold', echo=T, results="hide", message= F, warning=F}
plotData(dataset = "CLIMATE_LOCAL", site = "le_bray", automaticPanels = TRUE)
```

Or for TREE data
```{r, fig.show='hold',echo=T, results="hide", message= F, warning=F}
plotData(dataset = "TREE", site = "solling_beech", automaticPanels = TRUE)
```
Note that the argument *automaticPanels* defines whether the plot in a panel.

## Additional options

Besides, plotData supports the same options as getData. 

### forcingDataset, forcingCondition


```{r, echo=T, message= F, warning=F}
plotData(dataset ="CLIMATE_ISIMIP2B", site ="soro",
         forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
         automaticPanels = TRUE)
```

### variables


```{r, echo=T, fig.height=4,  message= F, warning=F}
plotData(dataset ="CLIMATE_ISIMIP2B",site ="soro", 
         forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
         variables = "p_mm")
```

### species

```{r, echo=T, message= F, warning=F}
plotData(dataset ="TREE", site ="hyytiala", species = "Pinus sylvestris",
         automaticPanels = TRUE)
```


### period

```{r, echo=T, message= F, warning=F}
plotData("CLIMATE_LOCAL", "soro", period = c("2011-01-01","2012-12-31"))
```

### quality

```{r, echo=T, message= F, warning=F}
plotData("CLIMATE_LOCAL", "soro", period = c("2011-01-01","2012-12-31"), 
         quality = 1, decreasing = FALSE)
```

```{r, echo=T, message= F, warning=F}
plotData("FLUX", "soro", period = c("2011-01-01","2012-12-31"), 
         quality = 0, decreasing = TRUE)
```

### aggregate

You can also aggregate data by defining *aggregate* and *FUN*. The data can be aggregated by day, month, year or date, and any statistic is supported, e.g. median.

```{r, fig.width=6, fig.height=4, warning = F,echo=T, results="hide", message= F}
plotData(dataset = "CLIMATE_ISIMIP2B", site ="soro",
         forcingDataset= "GFDLESM2M", forcingCondition="rcp2p6", 
         variables = "tmax_degC", period = c("2020-01-01", "2022-01-01"),
         aggregate = "month", FUN =median, automaticPanels = FALSE)

```




# The queryDB function: parsing your own queries

The package functions are designed to ease the access to the PROFOUND database by providing an R interface. However, if you feel confident with SQL syntax's, you could perform more advanced queries. The function *queryDB* allows to perform self-defined queries.

A basic query is *SELECT * FROM TABLE*, which will return all variables in the table. 
```{r, fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
overview <- queryDB("SELECT * FROM OVERVIEW")
tree <- queryDB("SELECT * FROM TREE")
```
More advanced queries include defining variables and conditions as the example below

```{r, fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL 
                   WHERE tmax_degC > 20 AND site == 'hyytiala' AND year == 2010")
```

Besides, the database contains several views with ready-made queries (check *browseData("DATASETS")*), so that the two statements below are equivalent

```{r, fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL 
                   WHERE site == 'hyytiala'")
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL_12")
```

This also works with tree species
```{r, fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT * FROM TREE 
                   WHERE species == 'Picea abies'")
myQuery <- queryDB("SELECT * FROM TREE_piab")
```

Hint: Check the last section for further information about SQL syntax's.
 

# The reportDB function

This function makes possible to create reates a site-by-site report of all avalaible data in the PROFOUND database. The summary is created with a rmarkdown document, which is rendered and saved as a html document. The report provdes an exhaustive description and visualization of the PROFOUND database. 

The function requires a path where to save the html document. If no path is specified, the working directory will be used as output directory.

```{r, eval=F, echo=T, results="hide", message= F, warning=F}
reportDB(outDir = "/home/database/")
```

Please note that creating the report it might take several minutes.

# Using the databse via sql 

To access the database you can use an SQLite database explorer such as [SQLiteStudio](http://sqlitestudio.pl/). Such software allows to visualize and download data through an interface, as well as writing SQL statements. Further information on SQLite characteristics can be found on the [SQLite website](https://www.sqlite.org/lang.html). Besides, there are plenty of resources about using SQL like [this one](http://www.w3schools.com/sql/default.asp) 

