---
title: "The PROFOUND R-package: ProfoundData"
author: "Ramiro Silveyra Gonzalez, Christopher Reyer, Florian Hartig, Friedrich Bonn, Klara Dolos"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Describes PROFOUND R-package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
--- 
  

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=10, warning=FALSE, cache = F)
```


```{r setup, echo=F, warning = F, message=F, results = 'hide'}
set.seed(1)
library(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
#panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
#set.alignment('left', row.names = 'right')
library(knitr)
library(plyr)

set.seed(1)


``` 



# ProfoundData R-package: using the PROFOUND database from R 
(modify)

The ProfoundData package serves as interface to access the [PROFOUND database](PROFOUNDdatabase.html). Three main functions are included to achieve this goal: a browseData function for exploring the database, a getData function for downloading data form the database and a plotData function to quickly inspect any variable of the data sets.

Currently, the functions allow to download data from one site and one data set at a time. For downloading data of more than one data set or of more than one site, the function should be called as many times as sites or data sets. Additionally, flags and other parameters can be passed to the functions.

The package includes other functions to provide utilities related to inspecting the data (`summarizeData()`) or write netcdf files (link to function)

This is just a preliminary version of both package and database. As user, please test the functions and provide feedback on what could be improved.


# First Steps

## Package requriments

Before installing the package, you should have the following packages installed:

* sqldf
* RSQLite
* zoo

## Database connection

After loading ProfoundData, the first step is to call the *setDB*. 
```{r, eval = T, message=FALSE, warning=FALSE}
library(ProfoundData)
```

This function will create a database object. It requires from you a **valid absolute path to the PROFOUND database**.

```{r, eval = T}
myDB <- path.expand("~/ownCloud/PROFOUND_Data/v0.1.13/ProfoundData.sqlite")
setDB(myDB)
```

To check the database connection you can use the use *getDB*, which returns the database path and also informs on the database version.

```{r, eval = T}
getDB()
```

# Exploring the database

## The browseData function

The browseData function allows to see what data is included in the database and for what sites the data is available. 

```{r, fig.width= 7}
overview <- browseData()
```
```{r , fig.width= 7}
knitr::kable(overview, align = "l")
```

Hint: If you set *collapse* to FALSE, you will get the full version of the overview table
```{r kable, fig.width= 7, eval = F}
overview <- browseData(collapse = F)
```


Besides returning the overview, browseData allows to check:

+  available data sets

```{r }
tables <- browseData(dataset = "DATASETS")
```

+  available variables for a given data set

```{r}
variables <- browseData(dataset = "CLIMATE_LOCAL", variables = T)
```

+  available sites for a given data set

```{r}
available <- browseData(dataset = "CLIMATE_LOCAL")
```


+  available data sets for a given site

```{r}
available <- browseData(site ="le_bray")
```

+  whether a data set is available for a specific site

```{r}
available <- browseData(site ="le_bray", dataset = "CLIMATE_LOCAL")
```

+  version history 

```{r}
version <- browseData(dataset = "VERSION")
```

+  metadata

```{r}
metadata <- browseData(dataset = "METADATA_DATASETS")
metadata <- browseData(dataset = "METADATA_CLIMATE_LOCAL")
```
Hint: You can also pass a site to check the specific site metadata
```{r}
metadata <- browseData(dataset = "METADATA_TREE", site = "solling_spruce")
```

+  source

```{r}
source <- browseData(dataset = "SOURCE")
```
Hint: You can also pass a site to check the specific site source
```{r}
source <- browseData(dataset = "SOURCE", site = "solling_spruce")
```

+  policy

```{r}
source <- browseData(dataset = "POLICY")
```
Hint: You can also pass a site to check the specific site source

```{r}
policy <- browseData(dataset = "POLICY", site = "solling_spruce")
```


## The summarizeData function

### Data overviews of the available data

The *summarizeData* allows to obtain data summaries and overview of the data, depending of **mode**. If you choose the mode overview, *summarizeDara* will return a data frame with the years and number of observations, the first as in the example below.

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_LOCAL", site = "bily_kriz", mode = "overview")
```
```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

The overview tables for ISIMIP datasets containing more fields (forcing datasets and conditions). 

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_ISIMIP2B", site = "bily_kriz", mode = "overview")
```
```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

This options is also availabe for TREE and STAND data sets.
```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "TREE", site = "solling_beech", mode = "overview")
```

```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

Please note that the TREE overview table is first summarized by years, pulling all trees together as a stand. Thus, you might find coincidences between the STAND and the TREE overviews.
```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "STAND", site = "solling_beech", mode = "overview")
```

```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

### Data summaries of the data

The other option of mode, **data**, sets the behaviour of *summarizeData* to to produce result summaries. 

#### TREE


```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "TREE", site = "bily_kriz")
```
```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```


#### CLIMATE

##### CLIMATE_LOCAL

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_LOCAL", site = "bily_kriz")
```
```{r, fig.show='hold', echo=T, message= F, warning=F}
kable(data, row.names = F)
```

##### CLIMATE_ISIMIP

For several forcing data sets and/or forcing conditions

```{r, echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_ISIMIP2B", site = "bily_kriz")
```
```{r, echo=T, message= F, warning=F}
kable(data, row.names = F)
```

For only one forcing data set and forcing conditions 
```{r, echo=T, message= F, warning=F}
data <-summarizeData(dataset = "CLIMATE_ISIMIP2B_IPSLCM5ALR_historical", site = "bily_kriz")
```
```{r, echo=T, message= F, warning=F}
kable(head(data), row.names = F)
```

#### FLUX


```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-summarizeData(dataset = "FLUX", site = "bily_kriz")
```
```{r, echo=T, message= F, warning=F}
kable(data, row.names = F)
```



# The getData function

The getData allows to download data from the PROFOUND database. The returned object will be a data frame or list of data frames, depending on the data that has been requested. 


## SITES

To obtain the site information, use getData to download the table SITES

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-getData(dataset = "SITES")
```
```{r, echo=T, message= F, warning=F}
names(data)
```

Optionally, you can specify a site

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-getData(dataset = "SITES", site = "soro")
```
```{r, echo=T, message= F, warning=F}
knitr::kable(data, row.names = F)
```

A text description can be obtained from **SITEDESCRIPTION**

```{r, fig.show='hold', echo=T, message= F, warning=F}
soro <-getData(dataset = "SITEDESCRIPTION", site = "soro")
```
```{r, echo=T, message= F, warning=F}
soro$description
```


## DATASETS

To download any data set, you have to provide a data set name and a site

```{r, fig.show='hold', echo=T, message= F, warning=F}
data <-getData( dataset = "CLIMATE_LOCAL", site = "soro")
```


### collapse

This option specifies whether the returned data should be data frame or a list of data frames. This argument is relevant when downloading data from SOIL or ISIMIP data sets.

For ISIMIP data sets, by setting collapse to FALSE you will obtain a list with data frames, named after the forcing data sets and conditions. Otherwise, the data will return in a unique data frame.

```{r, echo=T, message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2A", site = "soro", collapse = FALSE)
```
```{r, echo=T, message= F, warning=F}
names(data)
names(data[[1]])
```

We recommend to use unset collapse when downloading SOIL data, because in many cases SOIL is a collection of tables.(improve)

```{r, echo=T, message= F, warning=F}
data <- getData(dataset ="SOIL", site = "soro", collapse = FALSE)
```
```{r, echo=T, message= F, warning=F}
str(data, 1)
```
```{r, echo=T, message= F, warning=F}
names(data[[1]])
```


### forcingDataset, forcingCondition

When downloading ISIMIP data sets, regardless of whether it is climate, nitrogen deposition or CO2 data, you might want to select specific forcing data sets and/or conditions. With the function arguments *forcingDataset* and *forcingConditions* is possible, respectively.

```{r, echo=T, message= F, warning=F}
data <- getData( dataset ="CLIMATE_ISIMIP2B", site ="soro", forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6")
```
```{r, echo=T, message= F, warning=F}
knitr::kable(head(data), align = "l")

```

The abate getData call is actually equivalent to 
```{r, echo=T, message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B_GFDLESM2M_rcp2p6", site ="soro")
```
```{r, echo=T, message= F, warning=F}
knitr::kable(head(data), align = "l")
```


### variables

For any data set, it is possible to define the variable or variables to be downloaded

```{r, echo=T, message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B", site ="soro",
                forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
                variables = "p_mm")
```
```{r, echo=T,  message= F, warning=F}
knitr::kable(head(data),  align = "l")

```

### species
When downloading TREE or STAND data, you can select species with the argument *species*. It takes both full species name or the species id.
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="TREE", site ="hyytiala",
                species = "Pinus sylvestris")
```
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="TREE", site ="hyytiala",
                 species = "pisy")
```

### period
For time series data, it is possible to subset the data to a specific period.
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B", site ="soro",
                forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
                period = c("2006-01-01","2006-12-31"))
```
```{r, echo=T, message= F, warning=F}
range(data$date)
```

### quality

Some data sets, such as CLIMATE_LOCAL or FLUX, have quality flags to indicate how the data were obtained. Please be aware of the values of quality flags before using this option, the definition of the flags is available in the metadata. 

When passing a quality flag value, you should also define how the threshold should be applied. Values out the selected range will be dropped from the data based on your query.

```{r, echo=T, message= F, warning=F}
data <- getData(dataset = "CLIMATE_LOCAL", site = "soro",
                period = c("2011-01-01","2012-12-31"), 
                quality = 1, decreasing = FALSE)

```

```{r, echo=T, message= F, warning=F}
data <- getData(dataset = "FLUX", site = "soro",
                period = c("2011-01-01","2012-12-31"), 
                quality = 0, decreasing = TRUE)
```
# The plotData function
 
The plotData function performs a query to the database, downloads the data, applies the desired options (time period, quality flag) and shapes the final data into a time series for plotting. It requires a data set name and a site. Check  the help files to see what data sets are supported. 

## Normal plotting

A simple plot call looks like 

```{r, fig.show='hold', echo=T, results="hide", message= F, warning=F}
plotData(dataset = "CLIMATE_LOCAL", site = "le_bray", automaticPanels = TRUE)
```

Or for tree data
```{r, fig.show='hold',echo=T, results="hide", message= F, warning=F}
plotData(dataset = "TREE", site = "solling_beech", automaticPanels = TRUE)
```
Note that argument *automaticPanels* that defines whether the plot in a panel.

## Additional options

Besides, it support the same options as getData

### forcingDataset, forcingCondition


```{r, echo=T, message= F, warning=F}
plotData(dataset ="CLIMATE_ISIMIP2B", site ="soro",
         forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
         automaticPanels = TRUE)
```

### variables


```{r, echo=T, fig.height=4,  message= F, warning=F}
plotData(dataset ="CLIMATE_ISIMIP2B",site ="soro", 
         forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
         variables = "p_mm")
```

### species

```{r, echo=T, message= F, warning=F}
plotData(dataset ="TREE", site ="hyytiala", species = "Pinus sylvestris",
         automaticPanels = TRUE)
```


### period

```{r, echo=T, message= F, warning=F}
plotData("CLIMATE_LOCAL", "soro", period = c("2011-01-01","2012-12-31"))
```

### quality

```{r, echo=T, message= F, warning=F}
plotData("CLIMATE_LOCAL", "soro", period = c("2011-01-01","2012-12-31"), 
         quality = 1, decreasing = FALSE)
```

```{r, echo=T, message= F, warning=F}
plotData("FLUX", "soro", period = c("2011-01-01","2012-12-31"), 
         quality = 0, decreasing = TRUE)
```

### aggregate

You can also aggregate data by defining *aggregate* and *FUN*. The data can be aggregated by day, month, year or date, and any statistic is supported such as median

```{r, fig.width=6, fig.height=4, warning = F,echo=T, results="hide", message= F}
plotData(dataset = "CLIMATE_ISIMIP2B", site ="soro",
         forcingDataset= "GFDLESM2M", forcingCondition="rcp2p6", 
         variables = "tmax_degC", period = c("2020-01-01", "2022-01-01"),
         aggregate = "month", FUN =median, automaticPanels = FALSE)

```




# The queryDB function: parsing your own queries

The package functions are designed to ease the access to database by providing an R interface. However, if you feel confident with SQL syntax's, you are ready to perform more advanced queries. The function *queryDB*, allows to perform self-defined queries.

A basic query is *SELECT * FROM TABLE*, which will return all variables in the table. 
```{r, fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
overview <- queryDB("SELECT * FROM OVERVIEW")
tree <- queryDB("SELECT * FROM TREE")
```
More advanced queries include defining variables and conditions as the example below

```{r, fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL 
                   WHERE tmax_degC > 20 AND site == 'hyytiala' AND year == 2010")
```

Besides, the database contains several views with ready-made queries (check *browseData("DATASETS")*), so that the two statements below are equivalent

```{r, fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL 
                   WHERE site == 'hyytiala'")
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL_12")
```

This also works with tree species
```{r, fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT * FROM TREE 
                   WHERE species == 'Picea abies'")
myQuery <- queryDB("SELECT * FROM TREE_piab")
```

Hint: Check the last section for further information on SQL syntax's.
 
 
 
# Package structure

Both functions rely on several utility functions. Regardless of the requested data with the plotData or the getData, the following steps take place:

 1. Create and validate the SQL query (parseQuery.R). 
Creates a query item and validates it.

 2. Check the requested variables (parseVariables.R)
The requested variables are proven for spelling mistakes.
This is step is not included in the parseQuery function because it does not affect the query. All variables for the requested data set and site are download. Once download, only the selected variables will be displayed.


 3. Fetch the Query and download the data (fetchQuery.R)
If previous steps were successful, the data will be downloaded.


 4. Format data (formatData.R)
This function will transform the queried data into a data frame, or list of data frames depending on what was requested (see above examples). It relies on other functions such as



# Using the databse via sql 

To access the database you can use an SQLite database explorer such as [SQLiteStudio](http://sqlitestudio.pl/). Such software allows to visualize and download through an interface as well as writing SQL statements. Further information on SQLite characteristics can be found at the [SQLite website](https://www.sqlite.org/lang.html). Besides, there are plenty of resources about using SQL like [this one](http://www.w3schools.com/sql/default.asp) 

