---
title: "The PROFOUND R-package: ProfoundData"
author: "Ramiro Silveyra Gonzalez, Christopher Reyer, Florian Hartig, Friedrich Bonn, Klara Dolos"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Describes PROFOUND R-package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
--- 
  

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=10, warning=FALSE, cache = F)
```


```{r setup, echo=F, warning = F, message=F, results = 'hide'}
set.seed(1)
library(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
#panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
#set.alignment('left', row.names = 'right')

tableCaption <- local({
  tableid <- 0
  function(x=NULL){
    if(is.null(x)){
      return(tableid)
    }else{
      tableid <<- tableid + 1
      paste("Table ", tableid, ": ", x , sep = "")  
    }
  }
})

figCaption <- local({
  figureid <- 0
  function(x) {
    figureid <<- figureid + 1
    paste('Figure ', figureid, ': ', x, sep = '')
  }
})

subsectionNumbering <- local({
  sectionid <- 0
  function(x, y) {
    sectionid <<- sectionid + 1
    paste('## ', y, ".", sectionid, '. ', x, "\n\n", sep = '')
  }
})

library(knitr)
library(plyr)

set.seed(1)


``` 



# ProfoundData R-package: using the database from R 

This package serves as interface for users that want to access the PROFOUND data. Two main functions are included to achieve this goal: a getData function for downloading data form the database and a plotData function to quickly inspect any variable of the datasets.

Currently, the functions permit to download data from one location and one dataset at a time. For downloading data of more than one dataset or of more than one location, the function should be called as many times as locations or datasets. Additionally, flags and other parameters can be passed to the functions.

This is just a preliminary version of both package and database. As user, please test the functions and provide feedback on what could be improved.


# First Steps

Before installing the package, you should have the following packages available:

* sqldf
* RSQLite
* zoo


Once you have installed the package, it is time to use it. After loading ProfoundData, the first step is to call the *setDB*. This function will create a database object. It requires from you a **valid absolute path to the PROFOUNDData database**.  So lets start!

```{r, eval = T}
library(ProfoundData)
setDB("/home/ramiro/ownCloud/PROFOUND_Data/v0.1.13/ProfoundData.sqlite")
```

To check the database  you can use the use *getDB*, which returns the database path and also informs on the database version.
```{r, eval = F}
getDB()
```

# The browseData function: exploring the database


The browseData function allows you to see what data is included in the database and for what locations the data is available. Currently you will find that for several locations the data is not available.

```{r overviewKable, fig.width= 7}
overview <- browseData()
knitr::kable(overview,  align = "l")
```

Hint: If you set *collapse* to FALSE, you will get the full version of the overview table
```{r kable, fig.width= 7, eval = F}
overview <- browseData(collapse = F)
```


Besides returning the overview, browseData allows to check:

+  available datasets

```{r }
tables <- browseData(dataset = "DATASETS")
```

+  available variables for a given dataset

```{r}
variables <- browseData(dataset = "CLIMATE_LOCAL", variables = T)
```

+  available locations for a given dataset

```{r}
available <- browseData(dataset = "CLIMATE_LOCAL")
```


+  available datasets for a given location

```{r}
available <- browseData(location ="le_bray")
```

+  whether a dataset is available for a specific location

```{r}
available <- browseData(location ="le_bray", dataset = "CLIMATE_LOCAL")
```

+  version histoy 


```{r}
version <- browseData(dataset = "VERSION")
```


+  metadata

```{r}
metadata <- browseData(dataset = "METADATA_DATASETS")
metadata <- browseData(dataset = "METADATA_CLIMATE_LOCAL")
```
Hint: You can also pass a location to check the specific site metadata
```{r}
metadata <- browseData(dataset = "METADATA_TREE", location = "solling_spruce")
```

+  source

```{r}
source <- browseData(dataset = "SOURCE")
```
Hint: You can also pass a location to check the specific site source
```{r}
source <- browseData(dataset = "SOURCE", location = "solling_spruce")
```

+  policy

```{r}
source <- browseData(dataset = "POLICY")
```
Hint: You can also pass a location to check the specific site source

```{r}
policy <- browseData(dataset = "POLICY", location = "solling_spruce")
```


# The getData function

The getData allows to download data from the PROFOUND database. The returned object will be a dataframe or list of dataframes, depending on the data that has been requested. 


## SITES

To obtain the site information, use getData to download the table SITES

```{r, fig.show='hold', echo=T,  message= F, warning=F}
data <-getData(dataset =  "SITES")
names(data)
```

Optionally, you can specify a site

```{r, fig.show='hold', echo=T,  message= F, warning=F}
data <-getData(dataset =  "SITES", location = "soro")
knitr::kable(data, row.names = F)
```

A text description can be obtained from **SITEDESCRIPTION**

```{r, fig.show='hold', echo=T,  message= F, warning=F}
soro <-getData(dataset =  "SITEDESCRIPTION", location = "soro")
soro$description
```


## DATASETS

To download any dataset, you have to provide a dataset name and a location

```{r, fig.show='hold', echo=T,  message= F, warning=F}
data <-getData( dataset =  "CLIMATE_LOCAL", location = "soro")
```


### collapse

This option specifies wether the returned data should be data frame or a list of data frames. This argument is relevant when dowloading data from SOIL  or ISIMIP datasets.

For ISIMIP datasets, by setting collapse to FALSE youwill obtain a list with dataframes, named after the forcing datasets and conditions. Otherwise, the data will return in a unique data frame.

```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2A", location = "soro", collapse = FALSE)
names(data)
names(data[[1]])
```

We recommend always use unset collapse when downloading SOIL data, becuase in many cases SOIL is a collection of tables.

```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="SOIL", location = "soro", collapse = FALSE)
str(data, 1)
names(data[[1]])
```


### forcingDataset, forcingCondition

When downloading ISIMIP datasets, regaradless of whether it is climate, nitrogen deposition or CO2 data, you might want to select specific forcing datasets and/or conditions. With the function arguments *forcingDataset* and *forcingConditions* is possible, respectively.

```{r, echo=T,  message= F, warning=F}
data <- getData( dataset ="CLIMATE_ISIMIP2B", location ="soro", forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6")
knitr::kable(head(data),  align = "l")

```

The abote getData call is actually equivalent to 
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B_GFDLESM2M_rcp2p6", location ="soro")
knitr::kable(head(data),  align = "l")
```


### variables

For any dataset, it is possible to define the variable or variables to be downloaded

```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B", location ="soro", forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
                 variables = "p_mm")
knitr::kable(head(data),  align = "l")

```

### species

When downloading TREE or STAND data, you can select species with the argument *species*. It takes both full species name or the species id.
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="TREE", location ="hyytiala", species = "Pinus sylvestris")
```
```{r, echo=T,  message= F, warning=F}
data <- getData( location ="hyytiala", dataset ="TREE", species = "pisy")
```

### period
For time series data, it is possible to subset the data to a specific period.
```{r, echo=T,  message= F, warning=F}
data <- getData(dataset ="CLIMATE_ISIMIP2B", location ="soro", forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
                 period = c("2006-01-01","2006-12-31"))
range(data$date)
```

### quality

Some datasets, such as CLIMATE_LOCAL or FLUX, have qualitfy flags to indicate how the data were obtained. Please be aware of the values of quality flags before using this option, the definition of the flags is available in the metadata. 

When passing a quality flag value, you should also define how the thresholding should be applied. Values out the selected range will be dropped from the data based on your query.

```{r, echo=T,  message= F, warning=F}
data <- getData(dataset = "CLIMATE_LOCAL", location = "soro", period = c("2011-01-01","2012-12-31"), quality = 1, decreasing = FALSE)

```

```{r, echo=T,  message= F, warning=F}
data <- getData(dataset = "FLUX", location = "soro", period = c("2011-01-01","2012-12-31"), quality = 0, decreasing = TRUE)
```
# The plotData function
 
The plotData function performs a query to the database, downloads the data, applies the desired options (time period, quality flag) and shapes the final data into a time series for plotting. It requires a dataset name and a location. Check  the help files to see what datasets are supported. 

## Normal plotting

A simple plot call looks like 

```{r, fig.show='hold', echo=T, results="hide", message= F, warning=F}
plotData(dataset = "CLIMATE_LOCAL", location = "le_bray", automaticPanels = TRUE)
```

Or for tree data
```{r, fig.show='hold',echo=T, results="hide", message= F, warning=F}
plotData(dataset = "TREE", location = "solling_beech",   automaticPanels = TRUE)
```
Note that argument *automaticPanels* that defines whether the plot in a panel.

## Additional options

Besides, it support the same options than getData

-  **forcingDataset, forcingCondition**


```{r, echo=T,  message= F, warning=F}
plotData(dataset ="CLIMATE_ISIMIP2B", location ="soro", forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6", automaticPanels = TRUE)
```

- **variables**


```{r, echo=T, fig.height=4,  message= F, warning=F}
plotData(dataset ="CLIMATE_ISIMIP2B",location ="soro", forcingDataset="GFDLESM2M", forcingCondition ="rcp2p6",
                 variables = "p_mm")
```

- **species**

```{r, echo=T,  message= F, warning=F}
plotData(dataset ="TREE", location ="hyytiala", species = "Pinus sylvestris", automaticPanels = TRUE)
```


-  **period**

```{r, echo=T,  message= F, warning=F}
plotData("CLIMATE_LOCAL", "soro", period = c("2011-01-01","2012-12-31"))
```

- **quality**

```{r, echo=T,  message= F, warning=F}
plotData("CLIMATE_LOCAL", "soro", period = c("2011-01-01","2012-12-31"), quality = 1, decreasing = FALSE)
```

```{r, echo=T,  message= F, warning=F}
plotData("FLUX", "soro", period = c("2011-01-01","2012-12-31"), quality = 0, decreasing = TRUE)
```
- **aggregate**

You can also aggregate data by defining *aggregate* and *FUN*. The data can by aggregated by day, month,year or date, and any statistic is supported such as median

```{r,  fig.width=6, fig.height=4, warning = F,echo=T, results="hide", message= F}
plotData(dataset = "CLIMATE_ISIMIP2B", location ="soro",  forcingDataset= "GFDLESM2M",
           forcingCondition="rcp2p6", variables = "tmax_degC", period = c("2020-01-01", "2022-01-01"),
           aggregate = "month", FUN =median, automaticPanels = FALSE)

```



# The summarizeData function

The summarizeData allows to obtain data summaries from the PROFOUNDData database of climatic datasets and the TREE dataset.

## TREE


```{r, fig.show='hold', echo=T,  message= F, warning=F}
data <-summarizeData(dataset =  "TREE", location = "bily_kriz")
kable(data, row.names = F)
```

## CLIMATE

#### CLIMATE_LOCAL

```{r, fig.show='hold', echo=T,  message= F, warning=F}
data <-summarizeData(dataset =  "CLIMATE_LOCAL", location = "bily_kriz")
kable(data, row.names = F)
```

#### CLIMATE_ISIMIP

For several forcing datasets and/or forcing conditions

```{r, echo=T,  message= F, warning=F}
data <-summarizeData(dataset =  "CLIMATE_ISIMIP2B", location = "bily_kriz")
kable(data, row.names = F)
```

For only one forcing dataset and forcing conditions 
```{r, echo=T,  message= F, warning=F}
data <-summarizeData(dataset =  "CLIMATE_ISIMIP2B_IPSLCM5ALR_historical", location = "bily_kriz")
kable(head(data), row.names = F)
```


# The queryDB function: parsing your own queries

The package functions are designed to ease the access to database by providing an R interface. However, if you feel confident with SQL syntaxis, you are ready to perfom much advanced queries. The function *queryDB*, allows to perform self-defined queries, 

A basic query is *SELECT * FROM TABLE*, which will return all variables in the table. 
```{r,  fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
overview <- queryDB("SELECT * FROM OVERVIEW")
tree <- queryDB("SELECT * FROM TREE")
```
More advanced queries include defining variables and conditions as the example below

```{r,  fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL WHERE tmax_degC > 20 AND site == 'hyytiala' AND year == 2010")
```

Besides, the database contains several views with ready-made queries (check *browseData("DATASETS")*), so that the two statments below are equivalent

```{r,  fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL WHERE site == 'hyytiala'")
myQuery <- queryDB("SELECT date, tmax_degC FROM CLIMATE_LOCAL_12")
```

This also works with tree species
```{r,  fig.width=6, fig.height=4, echo=T, results="hide", message= F, warning=F}
myQuery <- queryDB("SELECT  * FROM TREE WHERE species == 'Picea abies'")
myQuery <- queryDB("SELECT  * FROM TREE_piab")
```

Hint: Check the last section for furhter information on SQL syntaxis.
 
 
 
# Package structure

Both functions rely on several utility functions. Regardless of the requested data with the plotData or the getData, the following steps take place:

 1. Create and validate the SQL query (parseQuery.R). 
Creates a query item and validates it.

 2. Check the requested variables (parseVariables.R)
The requested variables are proven for spelling mistakes.
This is step is not included in the parseQuery function because it does not affect the query. All variables for the requested dataset and location are download. Once download, only the selected variables will be displayed.


 3. Fetch the Query and download the data (fetchQuery.R)
If previous steps were succesfull, the data will be downloaded.


 4. Format data (formatData.R)
This function will transform the queried data into a dataframe, or list of dataframes depending on what was requested (see above examples). It relies on other functions such as



# Using the databse via sql 

To access the database  you  can use an SQLite database explorer such as [SQLiteStudio](http://sqlitestudio.pl/). Such software allows to visualize and download through an interface as well as writing SQL statements. Further information  on SQLite chracterisctics can be found at the [SQLite website](https://www.sqlite.org/lang.html). Besides, there are plenty of resources about using SQL like [this one](http://www.w3schools.com/sql/default.asp) 

